Tackling irreproducible enrichment analysis

### 2024-03-06 manuscript/manuscript.Rmd
intro

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) general lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.

To determine whether web based enrichment analysis is reproducible, we systematically examine the
reproducibilty of DAVID enrichment analysis in 20 published articles.
The results show that poor reporting and lack of preservation of web-tools is to blame for poor
reproducibility.
There is a need to bring enrichment analysis in line with bioinformatics best practices, so
we present a beginner-friendly protocol that guides researchers through the process of
literate scripting, version control, containerisation, documentation and code/data sharing.

@@

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring @Zeeberg2004-na.
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID version 6.8 (conducted March-May 2022) [@Sherman2022-hf].

Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically, this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
of Genes and Genomes (KEGG) pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

@@

```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility of enrichment analysis in a sample of 20 journal articles."}

score1 <- c("PMC6405693","PMC6425008","PMC6535219","PMC6539328","PMC6542760","PMC6561911","PMC6663624")
score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
  "PMC6582306","PMC6333352","PMC6526186","PMC6607402")
score3 <- c("PMC6349697","PMC6444048","PMC6587650", "PMC6368841")
l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)

@@

#grid()
text(0.7,1:7-0.5,labels=score1)
text(1.9,1:9-0.5,labels=score2)
text(3.1,1:4-0.5,labels=score3)

@@

>>>>

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) general lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results [@Sandve2013-lk].

To determine whether web based enrichment analysis is reproducible, here we systematically examine the
reproducibilty of DAVID enrichment analysis in 20 published articles, investigate root causes of
reproduction failure and describe some potential solutions.

@@

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring @Zeeberg2004-na.
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID v6.8 as at this time v6.7 was not available
(conducted March-May 2022) [@Sherman2022-hf].

Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically, this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results"
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
of Genes and Genomes (KEGG) pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

@@

```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility of enrichment analysis in a sample of 20 journal articles. Beside the PubMed Central identifier, the reported version of DAVID used in the original publication (blank if none given)."}

score1 <- c("PMC6405693 (v6.7)","PMC6425008 (v6.8)","PMC6535219","PMC6539328 (v6.8)",
  "PMC6542760","PMC6561911 (v6.8)","PMC6663624")
score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
  "PMC6582306","PMC6333352 (v6.8)","PMC6526186","PMC6607402 (v6.7)")
score3 <- c("PMC6349697 (v6.8)","PMC6444048","PMC6587650", "PMC6368841")
l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)

@@

#grid()
text(0.7,1:7-0.5,labels=score1,cex=0.85)
text(1.9,1:9-0.5,labels=score2,cex=0.85)
text(3.1,1:4-0.5,labels=score3,cex=0.85)

@@

The results show that poor reporting and lack of preservation of web-tools is to blame for poor
reproducibility.
There is a need to bring enrichment analysis in line with bioinformatics best practices, so
we present a beginner-friendly protocol that guides researchers through the process of
literate scripting, version control, containerisation, documentation and code/data sharing.
